---
title: "Case Study - Bike Sharing Company"
author: "Shubham Raj"
date: '2022-05-17'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Analysis of Cyclistic Rides for the Cyclistic Bike Share Company

### Load required packages
```{r}
library(tidyverse)  #helps wrangle data
library(lubridate)  #helps wrangle date attributes
library(ggplot2)  #helps visualize data
library(skimr) #for data summary statistics
library(janitor)
library(dplyr)
```

### Upload Divvy datasets from April 2021 to March 2022
```{r echo=TRUE}
apr21 <- read.csv('202104-divvy-tripdata.csv')
may21 <- read.csv('202105-divvy-tripdata.csv')
june21 <- read.csv('202106-divvy-tripdata.csv')
july21 <- read.csv('202107-divvy-tripdata.csv')
aug21 <- read.csv('202108-divvy-tripdata.csv')
sep21 <- read.csv('202109-divvy-tripdata.csv')
oct21 <- read.csv('202110-divvy-tripdata.csv')
nov21 <- read.csv('202111-divvy-tripdata.csv')
dec21 <- read.csv('202112-divvy-tripdata.csv')
jan22 <- read.csv('202201-divvy-tripdata.csv')
feb22 <- read.csv('202202-divvy-tripdata.csv')
mar22 <- read.csv('202203-divvy-tripdata.csv')
```

## Compare the columns of all dataset for inconsistency
```{r}
compare_df_cols(apr21,may21,june21,july21,aug21,sep21,oct21,nov21,dec21,
                jan22,feb22,mar22,return = "mismatch")
#we do not find any mismatch column data
```

### Combining all datasets into one dataset all_trips
```{r}
all_trips <- rbind(apr21,may21,june21,july21,aug21,sep21,oct21,nov21,dec21,
                   jan22,feb22,mar22)
```

### Inspect new dataset all_trips
```{r}
colnames(all_trips) #Get the list of column names
str(all_trips) #Get the list of column names and data type
```

### We don't need latitude and longitude data so we will remove it
```{r}
all_trips <- all_trips %>%
  select(-c(start_lat,start_lng,end_lat,end_lng))
```

### Inspect all_trips datasets after removing desired column
```{r}
colnames(all_trips)  #List of column names
dim(all_trips)  #Dimensions of the data frame?
str(all_trips)  #See list of columns and data types
summary(all_trips)  #Statistical summary of data. Mainly for numerics
table(all_trips$member_casual) #Number of each user type
```

### Add columns for date, day, month, year and day of the week
```{r}
all_trips$date <- as.Date(all_trips$started_at)
all_trips$day <- format(as.Date(all_trips$date), "%d")
all_trips$month <- format(as.Date(all_trips$date), "%m")
all_trips$year <- format(as.Date(all_trips$date), "%Y")
all_trips$day_of_week <- format(as.Date(all_trips$date), "%A")
```

### Add a "ride_length" calculation to all_trips (in seconds)
```{r}
all_trips <- all_trips %>%
  mutate(ride_length=difftime(ended_at,started_at))
```


### Inspect the structure of all columns
```{r}
str(all_trips)
```

### Convert "ride_length" to numeric so help us perform calculations on data
```{r}
all_trips$ride_length <- as.numeric(as.character(all_trips$ride_length))
is.numeric(all_trips$ride_length)
```
```{r count of ride_length<0}
count(all_trips[(all_trips$ride_length<0),])
```

### Remove Bad data
We will create a new version of the dataframe (v2) since we are removing data with ride_length<0
```{r}
all_tripsv2 <- all_trips[!(all_trips$ride_length <=0),]
```

### Conduct descriptive analysis on ride_length
```{r}
summary(all_tripsv2$ride_length) #summary stats of ride_length
```
### Compare members and casual riders
comparison of the mean, median, maximum and minimum ride_length by member_casual
```{r}
aggregate(all_tripsv2$ride_length ~ all_tripsv2$member_casual, FUN = mean)
aggregate(all_tripsv2$ride_length ~ all_tripsv2$member_casual, FUN = median)
aggregate(all_tripsv2$ride_length ~ all_tripsv2$member_casual, FUN = max)
aggregate(all_tripsv2$ride_length ~ all_tripsv2$member_casual, FUN = min)
```

### Average ride length for members and casul rider by the day of the week

```{r}
aggregate(all_tripsv2$ride_length ~ all_tripsv2$member_casual + all_tripsv2$day_of_week, FUN = mean)

```
```{r Order day_of_week}
#Days of week are out of order so we will fix this
all_tripsv2$day_of_week <- ordered(all_tripsv2$day_of_week, levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))
```

### Average ride length for members and casul rider by the day of the week
```{r}
aggregate(all_tripsv2$ride_length ~ all_tripsv2$member_casual + all_tripsv2$day_of_week,FUN = mean)
```

### Analyze ridership data on weekday by each type of rider
```{r}
all_tripsv2 %>%
   mutate(weekday = wday(started_at, label = TRUE)) %>%  #creates weekday field using wday()
  group_by(member_casual, weekday) %>%  #groups by usertype and weekday
  summarise(number_of_rides = n()				#calculates the number of rides and average duration 
  ,average_duration = mean(ride_length)) %>% #calculates the average duration
  arrange(member_casual, weekday) #sorting

```

### Visualize the number of rides per user type on week days
```{r}
all_tripsv2 %>%
  mutate(weekday=wday(started_at,label = TRUE)) %>%
  group_by(member_casual,weekday) %>%
  summarise(number_of_rides = n(),avg_duration = mean(ride_length)) %>%
  arrange(member_casual,weekday) %>%
  ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge")
```

### Visualize average ride duration by user type on week days
```{r}
all_tripsv2 %>% 
  group_by(member_casual, day_of_week) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, day_of_week)  %>% 
  ggplot(aes(x = day_of_week, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge")
```

### Visualize number o each user type
```{r}
ggplot(all_tripsv2,aes(x = member_casual, fill = member_casual))+ geom_bar()
```

### Export the data to CSV for further analysis
```{r}
write.csv(all_tripsv2, "cyclistic_trip_data.csv")
```
















